Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-06-08 manu_draft.Rmd
model predictions results figure added

# libraries
library(Jabbrev)
library(tidyverse)
library(wqtrends)
library(flextable)
library(officer)
library(ggmap)
library(sf)
library(ggsn)
library(USAboundaries)
library(hrbrthemes)

data(locs)
data(modprf)
data(seastrnd)

@@

Table \@ref(tab:modprftab), Table \@ref(tab:seastrndtab)

@@

p1 <- ggplot(toplo1, aes(x = yr, y = value, fill = Median, group = yr)) + 
  geom_violin(draw_quantiles = 0.5)  + 
  scale_fill_gradientn(colours = cls1) + 
  scale_y_log10() +
  theme_ipsum(base_family = 'serif') +
  theme(
    axis.title.x = element_blank()
  ) +
  labs(
    y = ylab, 
    subtitle = '(a) Annual summer/fall concentrations'
  )

p2 <- ggplot(toplo2, aes(x = mo, y = value, fill = yrcat)) + 
  geom_violin(draw_quantiles = 0.5)  + 
  scale_y_log10(ylab) +
  theme_ipsum(base_family = 'serif') +
  scale_fill_manual(values = cls2) +
  theme(
    legend.title = element_blank(), 
    axis.title.x = element_blank()
  ) +
  labs(
    y = ylab, 
    subtitle = '(b) Monthly concentrations by decade'
  )

p <- p1 + p2 + plot_layout(ncol = 1)
png('figs/obsdat.png', height = 8, width = 9, family = 'serif', units = 'in', res = 400)
print(p)
dev.off()
```
```{r obsdat, fig.cap = "Observed chlorophyll concentrations for all stations in Central and South SFE (18-36, Figure \\@ref(fig:sitemap)), with (a) annual summer/fall concentrations (Aug - Dec) and (b) monthly concentrations by decade."}
knitr::include_graphics('figs/obsdat.png')
```

@@

>>>>

# libraries
library(Jabbrev)
library(tidyverse)
library(wqtrends)
library(flextable)
library(officer)
library(ggmap)
library(sf)
library(ggsn)
library(USAboundaries)
library(hrbrthemes)
library(patchwork)
library(lubridate)
library(colorspace)

data(locs)
data(modprf)
data(seastrnd)
data(mods_chl18)
data(mods_chl24)
data(mods_chl27)
data(mods_chl32)

@@

Table \@ref(tab:modprftab), Table \@ref(tab:seastrndtab), Figure \@ref(fig:prddat)

@@

p1 <- ggplot(toplo1, aes(x = yr, y = value, fill = Median, group = yr)) + 
  geom_violin(draw_quantiles = 0.5)  + 
  scale_fill_gradientn(colours = cls1) + 
  scale_y_log10() +
  theme_minimal(base_family = 'serif') +
  theme(
    axis.title.x = element_blank()
  ) +
  labs(
    y = ylab, 
    subtitle = '(a) Annual summer/fall concentrations'
  )

p2 <- ggplot(toplo2, aes(x = mo, y = value, fill = yrcat)) + 
  geom_violin(draw_quantiles = 0.5)  + 
  scale_y_log10(ylab) +
  theme_minimal(base_family = 'serif') +
  scale_fill_manual(values = cls2) +
  theme(
    legend.title = element_blank(), 
    axis.title.x = element_blank()
  ) +
  labs(
    y = ylab, 
    subtitle = '(b) Monthly concentrations by decade'
  )

p <- p1 + p2 + plot_layout(ncol = 1)
png('figs/obsdat.png', height = 8, width = 8, family = 'serif', units = 'in', res = 400)
print(p)
dev.off()
```
```{r obsdat, fig.cap = "Observed chlorophyll concentrations for all stations in central and south San Francisco Estuary (18-36, Figure \\@ref(fig:sitemap)), with (a) annual summer/fall concentrations (Aug - Dec) and (b) monthly concentrations by decade."}
knitr::include_graphics('figs/obsdat.png')
```

@@

```{r, results = 'hide', message = F}
# get predictions to plot
prdplo <- bind_rows(mods_chl18, mods_chl24, mods_chl27, mods_chl32) %>% 
  mutate(prddat = pmap(list(data, modi), function(data, modi){

    prddat <- data.frame(
        dec_time = seq(min(data$dec_time), max(data$dec_time), length = 1000)
      ) %>%
      mutate(
        date = date_decimal(dec_time),
        date = as.Date(date),
        mo = month(date, label = TRUE),
        doy = yday(date),
        yr = year(date)
      )

    prd <- predict(modi, newdata = prddat) %>% 
      as.numeric

    annprd <- predict(modi, newdata = prddat, type = 'terms')
    int <- attr(annprd, 'constant')
    value <- rowSums(annprd) + int
    annvalue <- rowSums(annprd[, !grepl('doy', colnames(annprd)), drop = FALSE]) + int
    annvalue <- as.numeric(annvalue)

    trans <- unique(modi$trans)

    if(!inherits(trans, 'numeric'))
      stop('log used')

    prddat <- prddat %>% 
      mutate(
        chl = prd,
        chl = forecast::InvBoxCox(chl, trans), 
        annchl = forecast::InvBoxCox(annvalue, trans)
        )

    return(prddat)

  })) %>% 
  select(station, modi = model, prddat) %>% 
  unnest(prddat)

ylab <- expression(paste("Chlorophyll-a (", mu, "g ", L^-1, ")"))

toplo <- prdplo %>% 
  filter(yr >= 1991)

obsdat <- rawdat %>% 
  filter(station %in% c(18, 24, 27, 32)) %>% 
  filter(param == 'chl') %>% 
  filter(yr >= 1991) %>% 
  rename(chl = value)

colpal <- "PuBuGn"

p1 <- ggplot(toplo, aes(x = date, y = chl)) + 
  geom_point(data = obsdat, color = 'grey', size = 0.15) + 
  geom_line(size = 0.5) +
  geom_line(aes(y = annchl), color = 'tomato1', size = 0.5, alpha = 0.9) +
  theme_minimal(base_family = 'serif') + 
  theme(
    legend.position = 'right', 
    legend.title = element_blank(), 
    # axis.text.x= element_text(size = 8)
    strip.background = element_blank()
  ) + 
  facet_grid(modi ~ station) +
  scale_y_log10(limits = c(0.74, 83)) + 
  labs(x = 'Date', subtitle = '(a) Model predictions with annual estimates', y = ylab)

p2 <- ggplot(toplo, aes(x = doy, colour = yr, group = factor(yr))) + 
  geom_line(aes(y = chl)) + 
  theme_minimal(base_family = 'serif') + 
  theme(
    legend.position = 'right', 
    legend.title = element_blank(), 
    # axis.text.x= element_text(size = 8)
    strip.background = element_blank()
  ) + 
  scale_colour_continuous_sequential(palette = colpal, rev = T, begin = 0.2) +
  guides(colour = guide_colourbar(barheight = 10, barwidth = 1)) +
  facet_grid(modi ~ station) +
  scale_y_log10() + 
  labs(x = 'Day of year', subtitle = '(b) Model predictions by day of year', y = ylab)

p <- p1 + p2 + plot_layout(ncol = 1)

png('figs/prddat.png', height = 10, width = 9, family = 'serif', units = 'in', res = 400)
print(p)
dev.off()
```
```{r prddat, fig.cap = "GAM predictions for select stations from north to south (18, 24, 27, 32) for models in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix). The results show (a) predictions across the time series with annual estimates (red line) and (b) predictions by day of year with one smooth per year.  Observed data in (a) are shown with the gray points. Station locations are in Figure \\@ref(fig:sitemap)."}
knitr::include_graphics('figs/prddat.png')
```
